---
title: "Week 1: Introduction"
subtitle: "<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>"
author: Ian McCarthy | Emory University
date: Econ 470 & HLTH 470 #"`r format(Sys.time(), '%d %B %Y')`"
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts, custom.css] 
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---

<!-- Adjust some CSS code for font size and maintain R code font size -->
<style type="text/css">
.remark-slide-content {
    font-size: 30px;
    padding: 1em 2em 1em 2em;    
}
.remark-code, .remark-inline-code { 
    font-size: 20px;
}
</style>


<!-- Set R options for how code chunks are displayed at load packages -->
```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(knitr)
opts_chunk$set(
  fig.align="center",  
  fig.height=3, #fig.width=6,
  # out.width="748px", #out.length="520.75px",
  dpi=300, #fig.path='Figs/',
  cache=T#, echo=F, warning=F, message=F
  )

knitr::opts_hooks$set(fig.callout = function(options) {
  if(options$fig.callout) {
    options$echo = FALSE
  }
  options
})

library(tidyverse)
library(hrbrthemes)
library(scales)
```


# Table of contents

1. [Motivation](#welcome)

2. [Syllabus highlights](#syllabus)

3. [Getting started](#started)



<!-- New Section -->
---
class: inverse, center, middle
name: prologue

# Motivation

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>


---

# Health and Health Care

We've made *major* improvements in life expectancy (and many other measures of health) across the world

- Poverty reduction
- Technology development and innovation
- Technology diffusion and adoption
- Access to better services, including health care


---

# Evidence of better health

```{r life-exp, eval=FALSE}
library(gapminder)
gapminder %>%
  group_by(year) %>%
  summarize(lifeExp = median(lifeExp),
            gdpMed = median(gdpPercap)) %>%
  ggplot(aes(year,lifeExp)) + geom_line(alpha = 1/3) + theme_bw() +
    labs(x = "Year",
         y = "Life Expectancy (years)",
         title = "Median life expectancy across the world")

```
.plot-callout[
```{r life-exp-callout, ref.label="life-exp", fig.callout=TRUE}
```
]


---

# Evidence of better health

```{r life-exp-full-output, ref.label="life-exp", fig.callout=TRUE}
```


---

# GDP and Health

```{r lifexp-gdp, eval=FALSE}
library(gapminder)
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) + 
  geom_point(size = 1) + theme_bw() + scale_x_continuous(label = comma) +
  labs(x = "GDP Per Capita ($US)",
       y = "Life Expectancy (years)",
       title = "Life expectancy and GDP")
```
.plot-callout[
```{r lifexp-gdp-callout, ref.label="lifexp-gdp", fig.callout=TRUE}
```
]


---

# GDP and Health

```{r lifexp-full-output, ref.label="lifexp-gdp", fig.callout=TRUE}
```


---

# Spending and Health
```{r include=FALSE}
source("D:/CloudStation/Professional/Teaching Material/Emory/Econ 470 - Research in Health Economics and Policy/01-Introduction/DartmouthAtlas.R")
```


```{r dartmouth, eval=FALSE, warning=FALSE}
ggplot(data = (dartmouth.data %>% filter(Year==2015)), 
       mapping = aes(x = Expenditures, y = Total_Mortality)) + 
  geom_point(size = 1) + theme_bw() + scale_x_continuous(label = comma) +
  geom_smooth(method="lm", se=FALSE, color="blue", size=1/2) +
  labs(x = "Spending Per Capita ($US)",
       y = "Mortality Rate",
       title = "Mortality and Health Care Spending")
```
.plot-callout[
```{r dartmouth-callout, ref.label="dartmouth", fig.callout=TRUE, warning=FALSE}
```
]


---

# Spending and Health

```{r dartmouth-output, ref.label="dartmouth", fig.callout=TRUE, warning=FALSE}
```


---

# Spending and Health

- Does medical spending make us sicker?
- What else might explain this relationship?

.plot-callout[
```{r dartmouth-callout, ref.label="dartmouth", fig.callout=TRUE, warning=FALSE}
```
]


---

# Goals of this course.

1. Understand and implement selected methods for causal inference
--

2. Along the way...data management and version control with real data

--
3. Summarize, visualize, and explain research results


<!-- New Section -->
---
class: inverse, center, middle
name: syllabus

# Syllabus highlights

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

(Read the full document [here](Syllabus/Econ470-Syllabus.pdf).)

---

# Why this course?

---
class: inverse, center, middle

# Next lecture: Deep dive into Git(Hub).
<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>
