---
title: "Lecture 1: Introduction"
subtitle: "<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>"
author: Ian McCarthy | Emory University
date: EC 607 #"`r format(Sys.time(), '%d %B %Y')`"
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts] 
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(knitr)
opts_chunk$set(
  fig.align="center",  
  fig.height=4, #fig.width=6,
  # out.width="748px", #out.length="520.75px",
  dpi=300, #fig.path='Figs/',
  cache=T#, echo=F, warning=F, message=F
  )
library(tidyverse)
library(hrbrthemes)
```

# Table of contents

1. [Prologue](#prologue)

2. [Syllabus highlights](#syllabus)

3. [Getting started](#started)


---
class: inverse, center, middle
name: prologue

# Prologue

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---
class: inverse, center, middle
name: syllabus

# Syllabus highlights

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

(Read the full document [here](https://github.com/imccart/Econ-470/blob/master/Econ 470 Syllabus.pdf).)

---

# Why this course?

Fill in the gaps left by traditional econometrics and methods classes.

- Practical skills that tools that will benefit your dissertation and future career.
- Neglected skills like how to actually find datasets in the wild and clean them.

Data science skills are largely distinct from (and complementary to) the core 'metrics oeuvre familiar to economists.

- Data viz, cleaning and wrangling; programming; cloud computation; relational databases; machine learning; etc.

--

> *"In short, we will cover things that I wish someone had taught me when I was starting out in graduate
school."*

---

# Grading

| Component                           | Weight |
|:-|-:|
| 5 × homework assignments (15% each) | 75% |
| 2 × short presentations (5% each)		| 10% |
| 1 × peer-review						          | 10% |
|	In-class participation						  | 5%  |

- You can swap out one homework assignment for an (approved) final presentation of your own research.
- Short presentations summarize either a key lecture reading, or an (approved) software package/platform.
- We'll get to peer-review later in the course.

---
# Lecture outline

### Data science basics

- Introduction: Motivation, software installation, and data visualization
- Version control with Git(Hub)
- Learning to love the shell
- R language basics
- Data cleaning and wrangling with the "Tidyverse"
- Webscraping: (1) Server-side and CSS
- Webscraping: (2) Client-side and APIs

---
class: inverse, center, middle
name: started

# Getting started

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---

# Software installation and registration

1. Download [R](https://www.r-project.org/).

2. Download [RStudio](https://www.rstudio.com/products/rstudio/download/preview/).

3. Download [Git](https://git-scm.com/downloads).

4. Create an account on [GitHub](https://github.com/) and register for a student/educator [discount](https://education.github.com/discount_requests/new).

--

If you had trouble completing any of these steps, please raise your hand.
- My go-to place for installation guidance and troubleshooting is Jenny Bryan's http://happygitwithr.com.

---

# Checklist

☑ Do you have the most recent version of R?
  ```{r}
  version$version.string
  ```


☑ Do you have the most recent version of RStudio? (The [preview version](https://www.rstudio.com/products/rstudio/download/preview/) is fine.)
  ```{r eval=FALSE}
  RStudio.Version()$version
  ## Requires an interactive session but should return something like "[1] ‘1.1.463’"
  ```

☑ Have you updated all of your R packages? 
  ```{r eval=FALSE}
  update.packages(ask = FALSE, checkBuilt = TRUE)
  ```
  
---

# Checklist (cont.)

We will make sure that everything is working properly with your R and GitHub setup next lecture.

For the rest of today's lecture, I want to go over some very basic R concepts.

---
class: inverse, center, middle
name: r4ds

# R for data science
<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---

# Why R and RStudio? (cont.)
```{R, indeeddotcom, echo = F, fig.height = 6, fig.width = 9, dev = "svg"}
# The popularity data
pop_df <- 
  data.frame(
  lang = c("SQL", "Python", "R", "SAS", "Matlab", "SPSS", "Stata"),
  n_jobs = c(107130, 66976, 48772, 25644, 11464, 3717, 1624),
  free = c(T, T, T, F, F, F, F)
  )
## Plot it
pop_df %>% 
  mutate(lang = lang %>% factor(ordered = T)) %>%
  ggplot(aes(x = lang, y = n_jobs, fill = free)) +
  geom_col() +
  geom_hline(yintercept = 0) +
  aes(x = reorder(lang, -n_jobs), fill = reorder(free, -free)) +
  xlab("Statistical language") +
  scale_y_continuous(label = scales::comma) +
  ylab("Number of jobs") +
  labs(
    title = "Comparing statistical languages",
    subtitle = "Number of job postings on Indeed.com, 2019/01/06"
    ) +
  scale_fill_manual(
    "Free?",
    labels = c("True", "False"),
    values = c("#f92672", "darkslategray")
    ) +
  ggthemes::theme_pander(base_size = 17) +
  # theme_ipsum() +
  theme(legend.position = "bottom")
```

---

# Why R and RStudio? (cont.)

### Data science positivism

- Alongside Python, R has become the *de facto* language for data science.
  - See: [*The Impressive Growth of R*](https://stackoverflow.blog/2017/10/10/impressive-growth-r/), [*The Popularity of Data Science Software*](http://r4stats.com/articles/popularity/)
- Open-source (free!) with a global user-base spanning academia and industry.
  - "Do you want to be a profit source or a cost center?"

### Bridge to applied economics and other tools

- Already has all of the statistics and econometrics support, and is amazingly adaptable as a “glue” language to other
programming languages and APIs.
- The RStudio IDE and ecosystem allow for further, seemless integration.

### Path dependency

- It's also the language that I know best.
--

- (Learning multiple languages is a good idea, though.)

---
class: inverse, center, middle

# Next lecture: Deep dive into Git(Hub).
<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>
